---
title: "Handedness Advantage in Corner Outfield Defense"
author: '#TheBaseballAnalyst | @ProBaseballData'
date: "07-14-2025"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
load("rda/adjusted_of_data.rda")
load("rda/team_data.rda")
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

When a baseball is hit to left or right field, the side spin created through contact with the bat causes the ball to hook or tail toward the foul line. This phenomenon exists regardless of the handedness of the batter - i.e. a ball hit by a right-handed batter to left or right field will move in the direction of the foul line, and the same is true for light-handed batters. Given the movement of the ball in flight, a left-handed thrower, who has the glove on their right hand, could have an advantage over a right-handed thrower on balls hit in the air to left field. As the batted ball tails toward the foul line, the left fielder moves to their right to catch the ball, and with the glove on their right hand they might gain an advantage on the distance of the reach of their glove in that direction. The same is true for right-handed throwers playing right field; with the glove on their left hand, they could have an advantage as they move to their left to catch batted balls tailing toward the right field line.

For simplicity, left-handed throwing left fielders and right-handed throwing right fielders will be referred to as same-handed players, while right-handed left fielders and left-handed right fielders will be called opposite-handed players.

## Question

Do same-handed corner outfielders catch more balls hit in the air on average as compared to their opposite-handed counterparts?

## Data

Data used in this study is from the 2024 Major League Baseball season. Players who logged at least 100 innings in either corner outfield position are included. The total number of observations in the data set is
```{r, echo = FALSE, results = "asis"}
cat(nrow(adjusted_of_data))
```
, and the number of same-handed players in the data is
```{r, echo = FALSE, results = "asis"}
cat(adjusted_of_data %>%
  filter(Throws == 1) %>%
  nrow())
```
. A snapshot of the data set and descriptions of the variables are below:
```{r, echo = FALSE}
data <- adjusted_of_data %>%
  select(Name, Team, Pos, Throws_LR, Throws, PO.Inn, Sprint_Speed, teamFB.Inn) %>%
  rename("Position" = Pos, "Throws" = Throws_LR, "Throws (Indicator)" = Throws, "PO / Inn" = PO.Inn,
         "Sprint Speed" = Sprint_Speed, "Total Team Fly Balls / Inn" = teamFB.Inn)
knitr::kable(data[1:5, 1:8])
```

**Name** - First and last name of player. Note, players who logged 100 or more innings in both corner outfield positions appear twice in the data set.

**Team** - Team player played for. Note, players who appeared for multiple teams were removed from the data set. This was done to simplify the process of adjusting the data for Total Team Fly Balls / Inn during the regression analysis. 
  
**Position** - Position played by the player, either LF (left field) or RF (right field).

**Throws** - The hand the player throws with, either L (left) or R (right).

**Throws (Indicator)** - Indicator variable used to run regression analysis on the Throws variable. Same-handed players are assigned a value of 1 and opposite-handed throwers are given a value of 0.
  
**PO / Inn** - PO (putouts or catches made) divided by the number of defensive innings played at the position. The average across all players in the data set is
```{r, echo = FALSE, results = "asis"}
cat(round(sum(adjusted_of_data$PO.Inn)/nrow(adjusted_of_data), 4))
```
.

**Sprint Speed** - Statcast’s foot speed metric, defined as feet per second in a player’s fastest one-second window on individual plays. The average across all players in the data set is
```{r, echo = FALSE, results = "asis"}
cat(round(sum(adjusted_of_data$Sprint_Speed)/nrow(adjusted_of_data), 1))
```
.

**Total Team Fly Balls / Inn** - The total number of fly balls allowed by the team's pitching staff divided by the number of innings pitched, with home runs not included. Essentially, the number of balls hit in play per inning that could be catch-able fly balls. The average across all teams for the 2024 season is
```{r, echo = FALSE, results = "asis"}
cat(round(sum(team_data$cFB.IP)/nrow(team_data), 4))
```
.

## Analysis

Multiple linear regression was used to test if the independent variable Throws has an affect on the dependent variable PO / Inn. In essence, does a player's handedness affect the number of catches they make per inning in the corner outfield positions? There are also some additional factors to consider. Faster players should cover more ground and catch more fly balls than slower players. Also, players who play for a team that allows more fly balls per inning would have more chances and should record more putouts than players on teams whose pitching staffs allow fewer fly balls per inning. To adjust for both Sprint Speed and Total Team Fly Balls / Inn, both of these independent variables were added to the regression model. The results of the regression are as follows:

```{r, echo = FALSE}
PO_model <- lm(`PO / Inn` ~ `Throws (Indicator)` + `Sprint Speed` + `Total Team Fly Balls / Inn`, data = data)
summary(PO_model)
```

## Summary

The regression model shows a statistically significant positive relationship between same-handed corner outfielders and PO / Inn in the data set. However, only 20.65% of the variance in PO / Inn between players in the data is described by the independent variables included in the model (Throws, Sprint Speed, and Total Team Fly Balls / Inn), so there is uncertainty around the actual magnitude of the relationship between handedness and PO / Inn.

**Throws (Indicator)**

The estimated coefficient for the Throws variable is 
```{r, echo = FALSE, results = "asis"}
print(round(summary(PO_model)$coefficients[2,1], 4))
```
, and it is statistically significant from 0 meaning that there is a positive relationship between same-handed corner outfielders and PO / Inn. Since the indicator variable was used (1 for same-handed players and 0 for opposite-handed players), interpreted in context the coefficient provides the additional number of PO / Inn a same-handed player makes. This means that a same-handed corner outfielder has an advantage of 0.0183 PO / Inn on average over opposite-handed corner outfielders. Over the course of a full season with say 1000 innings played, that advantage would result in around 18 additional catches which is significant in terms of run value for the team. However, due to the low R-squared value of the regression, it is unlikely that the magnitude of the affect of throwing-handedness on PO / Inn is as high as the estimate produced by the model.

**Sprint Speed**
The estimated coefficient for the variable Sprint Speed is 
```{r, echo = FALSE, results = "asis"}
print(round(summary(PO_model)$coefficients[3,1], 4))
```
. Multiplying a player's Sprint Speed by .002 would indicate the number of catches per inning a player makes as a result of their Sprint Speed. So, if a player runs 27.9 ft/sec (the average of the players in the data set), then they gain .0558 catches per inning. Surprisingly, the model did not find the estimate of the coefficient on Sprint Speed to be statistically significant from 0, meaning there is no evidence in the data that faster runners make more catches per inning. This could be due to the fact that to play at least 100 major league innings in the corner outfield a player must be a competent runner to begin with, so all players in the data set are close enough in their speed for that variable to not be a significant factor. The average Sprint Speed of players in the data set is 27.9 ft/sec while the average across all major league players in 2024 is 27.0 ft/sec. In addition, other factors like positioning and jump may need to be adjusted for in order for Sprint Speed to gain statistical significance.

**Total Team Fly Balls / Inn**
The estimated coefficient for the variable Total Team Fly Balls / Inn is 
```{r, echo = FALSE, results = "asis"}
print(round(summary(PO_model)$coefficients[4,1], 4))
```
. This means if a team allows 0.9609 fly balls per innings (the major league average in 2024), then a corner outfielder would expect to make an additional 0.1821 catches per inning based solely on the number of non-home run fly balls allowed by the pitching staff of their team. Not surprisingly, the model found that the relationship between Total Team Fly Balls / Inn and PO / Inn is statistically significant from 0.

**R-Squared**
The model produced an R-squared value of 
```{r, echo = FALSE, results = "asis"}
print(round(summary(PO_model)$r.squared, 4))
```
, meaning only about 20.65% of the variance in PO / Inn between corner outfielders in the data set is due to the variables that were tested - Throws, Sprint Speed, and Total Team Fly Balls / Inn. With an R-squared value this low, there is likely a lot of noise in the data and that other factors beyond those studied here are contributing to PO / Inn for corner outfielders.

## Conclusion
The relationship between same-handed corner outfielders and PO / Inn is non-zero and positive for the observations in the data set, but due to the low R-squared value in the model it is unclear how impactful handedness is. Additional research is needed. Perhaps multiple seasons could be included to increase the size of the data set, and other variables should be considered such as outfielder positioning, jump, etc. Lastly, this was a quick analysis, but a more complete study of the topic might filter fly ball observations down to only those that require the outfielder to move in the direction of the foul line, or perhaps even just those fly balls in the "shadow" or "fringe" regions of the expected catch radius given the flight and trajectory of the batted ball and the starting position, jump, and foot speed of the player.

*Data Source: Fangraphs, Baseball Savant*